Density Matrix Susceptibility Tensors

### 2018-05-31 papers/1_dmatpol/paper-r1.tex
Change format of the paper, added abstract and summary sketch, added figure

\documentclass[aip,graphicx]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

@@

% mathematics
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{bm}
\usepackage{mathrsfs}
\usepackage{upgreek}
%\allowdisplaybreaks

@@

\draft % marks overfull lines with a black rule on the right

@@

\title{Density Matrix Susceptibility} %Title of paper

\author{Bartosz B{\l}asiak}
\email[]{blasiak.bartosz@gmail.com}
\homepage[]{https://www.polonez.pwr.edu.pl}
\thanks{Bulak}
\affiliation{Department of Physical and Quantum Chemistry, Faculty of Chemistry, Wroc{\l}aw University of Technology, Wybrze{\.z}e Wyspia{\'n}kiego 27, Wroc{\l}aw 50-370, Poland}

@@

\begin{abstract}
To be filled ...
\end{abstract}

@@

The electronic energy of a closed\hyp{}shell molecule $A$ is given by the time\hyp{}independent Schr{\"o}dinger equation,
%
\begin{equation}
 E^{(0)} = \tbraket{\Psi^{(0)}}{\mathscr{H}^{(0)}}{\Psi^{(0)}} \;,
\end{equation}
%
where $E$ stands for the total electronic energy, $\Psi$ the electronic wavefunction and $\mathscr{H}$
the Hamiltonian operator. In the above, the superscript $(0)$
denotes the reference or unperturbed state that is assumed to be known.
Under the influence of an external electrostatic perturbation, the total electronic energy
becomes
%
\begin{equation}
 E[{\bf F}({\bf r})] = \tbraket{\Psi}{\mathscr{H}^{(0)} + \mathscr{V}({\bf F}({\bf r}))}{\Psi}
\end{equation}
%
in which ${\bf F}({\bf r})$ is a non\hyp{}uniform electric field distribution.
The electric field\hyp{}induced energy change is an unknown functional 
of the associated change in the one\hyp{}electron density,
%
\begin{equation}\label{e:rho}
 \rho({\bf r}) = 2\sum_{\alpha\beta} D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
In Eq.~\eqref{e:rho} $\phi_\alpha({\bf r})$ is a basis funtion, which can be for instance 
an atomic orbital (AO) or a molecular orbital (MO), and $D_{\alpha\beta}$ is the corresponding
one\hyp{}particle density matrix.
If perturbation\hyp{}independent basis is chosen, the electric field\hyp{}induced change in the one\hyp{}particle density
is given by
%
\begin{equation}\label{e:rho}
 \delta \rho({\bf r}) = 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
Ideally, the electric field\hyp{}induced change in the ground state's one\hyp{}particle density 
follows its energy minimum of the perturbed state. In addition, all the perturbed 
excited state one\hyp{}electron densities must be excited state solutions of the Schr{\"o}dinger equation
with the exact energy\hyp{}minimized ground state.

@@

In the Hartree\hyp{}Fock theory, ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny has shown that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated differently.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{equation} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
    \right] } \;,
\end{equation}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This quarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the Paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

@@

  &=

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parametereized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w} = \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{equation}  \label{e:susceptibility-B}
 {\bf B}^{(i)}_{\alpha\beta} = \left\{
%
                               C_{\alpha i} {\bf b}_{\beta i} + C_{\beta i} {\bf b}_{\alpha i}
                               - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}
%
\right\}  \;.
\end{equation}
%
Hence, the 3D vector ${\bf B}^{(i)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i)}$
are unique properties of electronically unperturbed molecule.

@@

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarization.
Quadratic terms with respect to the electric field as well as linear terms with respect to
electric field gradient can be added to the formulation, which leads to
%
\begin{equation}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \left\{
                                      {\bf B}^{(i ; 10)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  
%                       + \sum_{ij}^M {\bf B}^{(ij; 20)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 20)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i)
                        +             {\bf B}^{(i ; 01)}_{\alpha\beta} : {\BM\nabla} \otimes {\bf F}({\bf r}_i) 
                        + \ldots \right\}
\end{equation}
%
%However, Eq.~\eqref{e:final-model} is an approximation because (i) it is based on the approximate
%relation in Eq.~\eqref{e:dmu-l-vector-mo-transform} and (ii) only total dipole moment (but not quadrupole
%and higher moments) is exactly
%reproduced by the current model. Due to the above reasons, the existence of ${\bf X}$ that
%reproduces the exact perturbed density matrix with vanishing error is not guaranteed. 
%Moreover, as it was pointed out before, ${\bf X}$ almost certainly depends on the external electric field. 
%Next section adresses the problem of finding the optimal ${\bf X}$ that is treated as an adjustable
%matrix.
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; RP)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field
and $P$-th order response with respect to the electric field gradient. Note that
${\bf B}^{(i; 10)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 20)}_{\alpha\beta}$
and ${\bf B}^{(i; 01)}_{\alpha\beta}$ are $(3\times 3)$ matrices. Also, the distributed expansion centres
(labeled as $i$) are generalized to any point in the molecule.

@@

Let  
$\left\{ {\bf F}^{(1)}({\bf r}), {\bf F}^{(2)}({\bf r}), \ldots, {\bf F}^{(N)}({\bf r}), \ldots \right\}$ 
be a set of $N_{\rm max}$ distinct and randomly sampled 
spatially non\hyp{}uniform distributions of the electric field. It is assumed that
the exact difference one\hyp{}particle density matrices (with respect to the unperturbed state)
defined as
%
\begin{equation}
 \delta \overline{\bf D}^{(N)} \equiv \overline{\bf D}^{(N)} - \overline{\bf D}^{(0)}
\end{equation}
%
are known for each sample (overline symbolizes the exact estimate).
%Let us assume that we a $\delta \overline{\bf D}^{(N)}$ 
%is the $N$th known (exact) difference one\hyp{}particle density matrix
%that was computed for the electric field distribution ${\bf F}^{(N)}({\bf r})$ chosen randomly.
Now, for each pair of the AO indices we construct the following 
parameterization of Eq.~\eqref{e:final-model.General}:
%
\begin{equation}\label{e:final-model.General.Parameters}
 \delta D^{(N)} = \sum_{i }^M \left\{ 
                              \sum_u^{x,y,z} S^{[1]}_{iu} F_{iu}^{(N)} 
%               + \sum_{ij}^M \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{jw} \right) F_{iu}^{(N)} F_{ju}^{(N)}
                +             \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{iw} \right) F_{iu}^{(N)} F_{iu}^{(N)}
                +             \sum_{uw}^{x,y,z} \left( S^{[3]}_{iu} + S^{[3]}_{iw}\right) F_{iu,iw}^{(N)}
                        + \ldots \right\}
\end{equation}
%
(the Greek subscripts were omitted here for notational simplicity).
The multiple parameter blocks 
$S^{[1]}$, $S^{[2]}$ and $S^{[3]}$ that are associated with electric field and its gradients
appear in the first power, allowing for linear least\hyp{}squares regression.
Note that $B_u^{(i,(10))} = S^{[1]}_{iu}$, $B_{uw}^{(i,(20))} = S^{[2]}_{iu} + S^{[2]}_{iw}$
and $B_{uw}^{(i,(01))} = S^{[3]}_{iu} + S^{[3]}_{iw}$. The square bracket superscripts 
denote the block of the parameter space. We chose symmetric forms of ${\bf B}^{(i;20)}$
and ${\bf B}^{(i;01)}$ (antisymmetric form could be constructed as well by taking the difference
of the parameters from block 2 and 3, respectively). $F_{iu}^{(N)}$ and $F_{iu,iw}^{(N)}$ 
stand for the electric field $\left[ {\bf F}({\bf r}_i)\right]_u$ and the electric field
gradient $\left[ {\BM \nabla} \otimes {\bf F}({\bf r}_i) \right]_{uw}$, respectively.

@@

In order to validate all the models discussed in this Contribution,
the electric field\hyp{}induced interaction energy, 
the dipole moment and the quadrupole moment were analysed in detail.
The electric field\hyp{}induced interaction energy at the HF level, $\delta E$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E[{\bf F}] = 
                     \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} \right) \right\}}
                   + \Tr{\left\{ \left( {\bf D}^{(0)} + \delta {\bf D} \right) \cdot
                                \left( 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} 
                   + E_{\rm nuc}({\bf F}) \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarization of the electron density distribution, respectively. $E_{\rm nuc}({\bf F})$ denotes the interaction energy
due to the nuclei.
In order to compare the performance of the distributed polarizability approach,
the approximate interaction energy was also evaluated by using the expression:
%
\begin{equation}\label{e:dE-dpol}
 \delta E[{\bf F}] \approx 2 \Tr{\left\{ {\bf D}^{(0)} {\bf V}\right\} } 
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i)
                   + E_{\rm nuc}({\bf F}) \;,
\end{equation}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic polarizability tensor associated with the
$i$th localized molecular orbital (LMO) centroid, ${\bf r}_i = \tbraket{\phi_i}{\bf r}{\phi_i}$.
The distributed polarizabilities were computed by utilizing the coupled\hyp{}perturbed Hartree\hyp{}Fock
method. Orbital localization was performed by using the Boys method. Note that in Eq.~\eqref{e:dE-dpol}, 
multipole approximation is introduced only for the part associated with the electric field\hyp{}induced
change in the one\hyp{}particle density matrix, i.e., the polarization of the electron density distribution.

>>>>

\documentclass[aip,amsmath,amssymb,reprint]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

% mathematics
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsbsy}
%\usepackage{bm}
\usepackage{mathrsfs}
\usepackage{upgreek}
%\allowdisplaybreaks

@@

%\draft % marks overfull lines with a black rule on the right

@@

\preprint{AIP/123-DMS}

\title{Density Matrix Susceptibility Tensors} %Title of paper

\author{Bartosz B{\l}asiak}
\email[]{blasiak.bartosz@gmail.com}
\homepage[]{https://www.polonez.pwr.edu.pl}
%\thanks{Bulak}
\affiliation{Department of Physical and Quantum Chemistry, Faculty of Chemistry, Wroc{\l}aw University of Technology, Wybrze{\.z}e Wyspia{\'n}kiego 27, Wroc{\l}aw 50-370, Poland}

@@

\begin{abstract}

%Considering the solvent effect on the correlated electronic wavefunctions, especially
%those that describe the electronically excited state, is a great challenge of Quantum Chemistry.

In this work, the electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibility tensors
interacting with electric fields. The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The interaction\hyp{}induced energies
dipole moments and quadrupole moments
are quantitatively described by the present theory in weak and moderate electric fields.
The possibility of extension of the proposed model with high\hyp{}order susceptibilities
opens new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
\end{abstract}

@@

The electronic energy of a closed\hyp{}shell molecule $A$ is given by the time\hyp{}independent Schr{\"o}dinger equation,
%
\begin{equation}
 E^{(0)} = \tbraket{\Psi^{(0)}}{\mathscr{H}^{(0)}}{\Psi^{(0)}} \;,
\end{equation}
%
where $E$ stands for the total electronic energy, $\Psi$ the electronic wavefunction and $\mathscr{H}$
the Hamiltonian operator. In the above, the superscript $(0)$
denotes the reference or unperturbed state that is assumed to be known.
Under the influence of an external electrostatic perturbation, the total electronic energy
becomes
%
\begin{equation}
 E[{\bf F}({\bf r})] = \tbraket{\Psi}{\mathscr{H}^{(0)} + \mathscr{V}({\bf F}({\bf r}))}{\Psi}
\end{equation}
%
in which ${\bf F}({\bf r})$ is a non\hyp{}uniform electric field distribution.
The electric field\hyp{}induced energy change is an unknown functional 
of the associated change in the one\hyp{}electron density,
%
\begin{equation}\label{e:rho}
 \rho({\bf r}) = 2\sum_{\alpha\beta} D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
In Eq.~\eqref{e:rho} $\phi_\alpha({\bf r})$ is a basis funtion, which can be for instance 
an atomic orbital (AO) or a molecular orbital (MO), and $D_{\alpha\beta}$ is the corresponding
one\hyp{}particle density matrix.
If perturbation\hyp{}independent basis is chosen, the electric field\hyp{}induced change in the one\hyp{}particle density
is given by
%
\begin{equation}\label{e:rho-1}
 \delta \rho({\bf r}) = 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
Ideally, the electric field\hyp{}induced change in the ground state's one\hyp{}particle density 
follows its energy minimum of the perturbed state. In addition, all the perturbed 
excited state one\hyp{}electron densities must be excited state solutions of the Schr{\"o}dinger equation
with the exact energy\hyp{}minimized ground state.

@@

In the Hartree\hyp{}Fock theory, ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny has shown that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated differently.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This quarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the Paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

@@

  & \\ \nonumber =

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parametereized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w} = \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i)}_{\alpha\beta} = \Big\{
                               C_{\alpha i} {\bf b}_{\beta i} + C_{\beta i} {\bf b}_{\alpha i} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}
%
\Big\}  \;.
\end{multline}
%
Hence, the 3D vector ${\bf B}^{(i)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i)}$
are unique properties of electronically unperturbed molecule.

@@

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarization.
Quadratic terms with respect to the electric field as well as linear terms with respect to
electric field gradient can be added to the formulation, which leads to
%
\begin{multline}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \Big\{
                                      {\bf B}^{(i ; 10)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  
%                       + \sum_{ij}^M {\bf B}^{(ij; 20)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 20)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \\
                        +             {\bf B}^{(i ; 01)}_{\alpha\beta} : {\BM\nabla} \otimes {\bf F}({\bf r}_i) 
                        + \ldots \Big\}
\end{multline}
%
%However, Eq.~\eqref{e:final-model} is an approximation because (i) it is based on the approximate
%relation in Eq.~\eqref{e:dmu-l-vector-mo-transform} and (ii) only total dipole moment (but not quadrupole
%and higher moments) is exactly
%reproduced by the current model. Due to the above reasons, the existence of ${\bf X}$ that
%reproduces the exact perturbed density matrix with vanishing error is not guaranteed. 
%Moreover, as it was pointed out before, ${\bf X}$ almost certainly depends on the external electric field. 
%Next section adresses the problem of finding the optimal ${\bf X}$ that is treated as an adjustable
%matrix.
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; RP)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field
and $P$-th order response with respect to the electric field gradient. Note that
${\bf B}^{(i; 10)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 20)}_{\alpha\beta}$
and ${\bf B}^{(i; 01)}_{\alpha\beta}$ are $(3\times 3)$ matrices. Also, the distributed expansion centres
(labeled as $i$) are generalized to any point in the molecule.

@@

Let  
$\left\{ {\bf F}^{(1)}({\bf r}), {\bf F}^{(2)}({\bf r}), \ldots, {\bf F}^{(N)}({\bf r}), \ldots \right\}$ 
be a set of $N_{\rm max}$ distinct and randomly sampled 
spatially non\hyp{}uniform distributions of the electric field. It is assumed that
the exact difference one\hyp{}particle density matrices (with respect to the unperturbed state)
defined as
%
\begin{equation}
 \delta \overline{\bf D}^{(N)} \equiv \overline{\bf D}^{(N)} - \overline{\bf D}^{(0)}
\end{equation}
%
are known for each sample (overline symbolizes the exact estimate).
%Let us assume that we a $\delta \overline{\bf D}^{(N)}$ 
%is the $N$th known (exact) difference one\hyp{}particle density matrix
%that was computed for the electric field distribution ${\bf F}^{(N)}({\bf r})$ chosen randomly.
Now, for each pair of the AO indices we construct the following 
parameterization of Eq.~\eqref{e:final-model.General}:
%
\begin{multline}\label{e:final-model.General.Parameters}
 \delta D^{(N)} = \sum_{i }^M \Big\{ 
                              \sum_u^{x,y,z} S^{[1]}_{iu} F_{iu}^{(N)} \\
%               + \sum_{ij}^M \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{jw} \right) F_{iu}^{(N)} F_{jw}^{(N)}
                +             \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{iw} \right) F_{iu}^{(N)} F_{iw}^{(N)} \\
                +             \sum_{uw}^{x,y,z} \left( S^{[3]}_{iu} + S^{[3]}_{iw}\right) F_{iu,iw}^{(N)}
                        + \ldots \Big\}
\end{multline}
%
(the Greek subscripts were omitted here for notational simplicity).
The multiple parameter blocks 
$S^{[1]}$, $S^{[2]}$ and $S^{[3]}$ that are associated with electric field and its gradients
appear in the first power, allowing for linear least\hyp{}squares regression.
Note that $B_u^{(i,(10))} = S^{[1]}_{iu}$, $B_{uw}^{(i,(20))} = S^{[2]}_{iu} + S^{[2]}_{iw}$
and $B_{uw}^{(i,(01))} = S^{[3]}_{iu} + S^{[3]}_{iw}$. The square bracket superscripts 
denote the block of the parameter space. We chose symmetric forms of ${\bf B}^{(i;20)}$
and ${\bf B}^{(i;01)}$ (antisymmetric form could be constructed as well by taking the difference
of the parameters from block 2 and 3, respectively). $F_{iu}^{(N)}$ and $F_{iu,iw}^{(N)}$ 
stand for the electric field $\left[ {\bf F}({\bf r}_i)\right]_u$ and the electric field
gradient $\left[ {\BM \nabla} \otimes {\bf F}({\bf r}_i) \right]_{uw}$, respectively.

@@

In order to validate all the models discussed in this Contribution,
the electric field\hyp{}induced interaction energy, 
the dipole moment and the quadrupole moment were analysed in detail.
The electric field\hyp{}induced interaction energy at the HF level, $\delta E$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E[{\bf F}] = E_{\rm nuc}({\bf F}) +
                     \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} \right) \right\}} \\
                   + \Tr{\left\{ \left( {\bf D}^{(0)} + \delta {\bf D} \right) \cdot
                                \left( 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarization of the electron density distribution, respectively. $E_{\rm nuc}({\bf F})$ denotes the interaction energy
due to the nuclei.
In order to compare the performance of the distributed polarizability approach,
the approximate interaction energy was also evaluated by using the expression:
%
\begin{multline}\label{e:dE-dpol}
 \delta E[{\bf F}] \approx E_{\rm nuc}({\bf F}) + 
                     2 \Tr{\left\{ {\bf D}^{(0)} {\bf V}\right\} }  \\
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{multline}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic polarizability tensor associated with the
$i$th localized molecular orbital (LMO) centroid, ${\bf r}_i = \tbraket{\phi_i}{\bf r}{\phi_i}$.
The distributed polarizabilities were computed by utilizing the coupled\hyp{}perturbed Hartree\hyp{}Fock
method. Orbital localization was performed by using the Boys method. Note that in Eq.~\eqref{e:dE-dpol}, 
multipole approximation is introduced only for the part associated with the electric field\hyp{}induced
change in the one\hyp{}particle density matrix, i.e., the polarization of the electron density distribution.
